---
title: "Multiple linear regression"
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

## Grading the professor

Many college courses conclude by giving students the opportunity to evaluate 
the course and the instructor anonymously. However, the use of these student 
evaluations as an indicator of course quality and teaching effectiveness is 
often criticized because these measures may reflect the influence of 
non-teaching related characteristics, such as the physical appearance of the 
instructor. The article titled, "Beauty in the classroom: instructors' 
pulchritude and putative pedagogical productivity" (Hamermesh and Parker, 2005) 
found that instructors who are viewed to be better looking receive higher 
instructional ratings. (Daniel S. Hamermesh, Amy Parker, Beauty in the 
classroom: instructors pulchritude and  putative pedagogical productivity, 
*Economics of Education Review*, Volume 24, Issue 4, August 2005, Pages 369-376, 
ISSN 0272-7757, 10.1016/j.econedurev.2004.07.013. [http://www.sciencedirect.com/science/article/pii/S0272775704001165](http://www.sciencedirect.com/science/article/pii/S0272775704001165).)

In this lab we will analyze the data from this study in order to learn what goes 
into a positive professor evaluation.

## The data

The data were gathered from end of semester student evaluations for a large 
sample of professors from the University of Texas at Austin. In addition, six 
students rated the professors' physical appearance. (This is aslightly modified 
version of the original data set that was released as part of the replication 
data for *Data Analysis Using Regression and Multilevel/Hierarchical Models* 
(Gelman and Hill, 2007).) The result is a data frame where each row contains a 
different course and columns represent variables about the courses and professors.

```{r load-data, eval=FALSE}
load("more/evals.RData")
```

variable         | description
---------------- | -----------
`score`          | average professor evaluation score: (1) very unsatisfactory - (5) excellent.
`rank`           | rank of professor: teaching, tenure track, tenured.
`ethnicity`      | ethnicity of professor: not minority, minority.
`gender`         | gender of professor: female, male.
`language`       | language of school where professor received education: english or non-english.
`age`            | age of professor.
`cls_perc_eval`  | percent of students in class who completed evaluation.
`cls_did_eval`   | number of students in class who completed evaluation.
`cls_students`   | total number of students in class.
`cls_level`      | class level: lower, upper.
`cls_profs`      | number of professors teaching sections in course in sample: single, multiple.
`cls_credits`    | number of credits of class: one credit (lab, PE, etc.), multi credit.
`bty_f1lower`    | beauty rating of professor from lower level female: (1) lowest - (10) highest.
`bty_f1upper`    | beauty rating of professor from upper level female: (1) lowest - (10) highest.
`bty_f2upper`    | beauty rating of professor from second upper level female: (1) lowest - (10) highest.
`bty_m1lower`    | beauty rating of professor from lower level male: (1) lowest - (10) highest.
`bty_m1upper`    | beauty rating of professor from upper level male: (1) lowest - (10) highest.
`bty_m2upper`    | beauty rating of professor from second upper level male: (1) lowest - (10) highest.
`bty_avg`        | average beauty rating of professor.
`pic_outfit`     | outfit of professor in picture: not formal, formal.
`pic_color`      | color of professor's picture: color, black & white.

## Exploring the data

1.  Is this an observational study or an experiment? The original research
    question posed in the paper is whether beauty leads directly to the
    differences in course evaluations. Given the study design, is it possible to
    answer this question as it is phrase. If not, rephrase the question

This is an observational study since there are no experimental conditions (e.g. no random assignment of subjects to intervention/non-intervention group.) No, it is not possible to directly answer this question. The question should be something like "are a teacher's looks associated with his course evaluation score."

2.  Describe the distribution of `score`. Is the distribution skewed? What does 
    that tell you about how students rate courses? Is this what you expected to 
    see? Why, or why not?

```{r hist-score,eval = FALSE}
hist(evals$score)
```

The distribution is unimodal but not symmetric.  It has a left-sided skew. I probably would have expected a mode on the high end (assuming students are generally sympethetic to their professors), but I would not have expected the mode to be 4.5 - that seems to be very high. The skew makes sense, as there is some range of low scores for students who did not like their professors.

3.  Excluding `score`, select two other variables and describe their relationship 
    using an appropriate visualization (scatterplot, side-by-side boxplots, or 
    mosaic plot).

```{r plot-others,eval = FALSE}
boxplot(evals$bty_f1lower,evals$bty_m1lower)
```


## Simple linear regression

The fundamental phenomenon suggested by the study is that better looking teachers
are evaluated more favorably. Let's create a scatterplot to see if this appears 
to be the case:

```{r scatter-score-bty_avg, eval = FALSE}
plot(evals$score ~ evals$bty_avg)
```

Before we draw conclusions about the trend, compare the number of observations 
in the data frame with the approximate number of points on the scatterplot. 
Is anything awry?

There appear to be far fewer observations in this scatterplot than there are overall observations in the data frame.

4.  Replot the scatterplot, but this time use the function `jitter()` on the
    $y$- or the $x$-coordinate. (Use `?jitter` to learn more.) What was
    misleading about the initial scatterplot?

```{r scatter-jitter, eval = FALSE}
plot(jitter(evals$score,3) ~ evals$bty_avg)
```

```{r scatter-jitter2, eval = FALSE}
plot(evals$score ~ jitter(evals$bty_avg, 3))
```

The jitter function adds some noise to the scatterplot to distinguish the number of observations, and make the relationship between x and y variables clearer. The original scatterplot did not clearly indicate the density of observations at high levels of the score variable - especially for the lower beauty scores.

5.  Let's see if the apparent trend in the plot is something more than
    natural variation. Fit a linear model called `m_bty` to predict average
    professor score by average beauty rating and add the line to your plot
    using `abline(m_bty)`. Write out the equation for the linear model and
    interpret the slope. Is average beauty score a statistically significant
    predictor? Does it appear to be a practically significant predictor?

```{r scatter-lm-line, eval = FALSE}
m_bty <- lm(evals$score ~ evals$bty_avg)
abline(
```

score = 3.88 + 0.067(bty_avg)

6.  Use residual plots to evaluate whether the conditions of least squares
    regression are reasonable. Provide plots and comments for each one (see
    the Simple Regression Lab for a reminder of how to make these).
    
```{r plotss-score-bty-squares, eval=FALSE}
plot_ss(x = evals$bty_avg, y = evals$score, showSquares = TRUE)
```

I tried several of these. As I got closer to the model noted above, the least squares score seems to have gone down.

Coefficients:
(Intercept)            x  
      2.997        0.219  

Sum of Squares:  177.315

Coefficients:
(Intercept)            x  
    3.76625      0.05112  

Sum of Squares:  147.568

Coefficients:
(Intercept)            x  
    3.63136      0.07505  

Sum of Squares:  152.716

Coefficients:
(Intercept)            x  
    3.84453      0.04264  

Sum of Squares:  141.804

(Intercept)            x  
    3.90799      0.08489  

Sum of Squares:  137.655

Coefficients:
(Intercept)            x  
    3.72087      0.08253  

Sum of Squares:  135.828

## Multiple linear regression

The data set contains several variables on the beauty score of the professor: 
individual ratings from each of the six students who were asked to score the 
physical appearance of the professors and the average of these six scores. Let's 
take a look at the relationship between one of these scores and the average 
beauty score.

```{r bty-rel, eval = FALSE}
plot(evals$bty_avg ~ evals$bty_f1lower)
cor(evals$bty_avg, evals$bty_f1lower)
```

Correlation is 0.844

As expected the relationship is quite strong - after all, the average score is 
calculated using the individual scores. We can actually take a look at the 
relationships between all beauty variables (columns 13 through 19) using the 
following command:

```{r bty-rels, eval = FALSE}
plot(evals[,13:19])
```

These variables are collinear (correlated), and adding more than one of these 
variables to the model would not add much value to the model. In this 
application and with these highly-correlated predictors, it is reasonable to use
the average beauty score as the single representative of these variables.

In order to see if beauty is still a significant predictor of professor score 
after we've accounted for the gender of the professor, we can add the gender 
term into the model.

```{r scatter-score-bty_avg_gender, eval = FALSE}
m_bty_gen <- lm(score ~ bty_avg + gender, data = evals)
summary(m_bty_gen)
```

7.  P-values and parameter estimates should only be trusted if the
    conditions for the regression are reasonable. Verify that the conditions
    for this model are reasonable using diagnostic plots.

```{r residuals, eval=FALSE}
plot(m_bty_gen$residuals ~ evals$bty_avg)
plot(m_bty_gen$residuals ~ evals$gender)
hist(m_bty_gen$residuals)
qqnorm(m_bty_gen$residuals)
qqline(m_bty_gen$residuals)
```

Residuals appear to be relatively consistent across different values of each predictor variable. The distribution of the residuals is unimodal, but does not appear exactly normal. That said the QQ plot/line looks relatively normal (with some deviation at the higher quartiles.)

8.  Is `bty_avg` still a significant predictor of `score`? Has the addition
    of `gender` to the model changed the parameter estimate for `bty_avg`?

Yes, `bty_avg` is still a significant predictor of score. The p value is very low, suggesting a significant relationship between the variables. The coefficient for `bty_avg` has gone up slightly from 0.067 to 0.074.

Note that the estimate for `gender` is now called `gendermale`. You'll see this 
name change whenever you introduce a categorical variable. The reason is that R 
recodes `gender` from having the values of `female` and `male` to being an 
indicator variable called `gendermale` that takes a value of $0$ for females and
a value of $1$ for males. (Such variables are often referred to as "dummy" 
variables.)

As a result, for females, the parameter estimate is multiplied by zero, leaving 
the intercept and slope form familiar from simple regression.

\[
  \begin{aligned}
\widehat{score} &= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg + \hat{\beta}_2 \times (0) \\
&= \hat{\beta}_0 + \hat{\beta}_1 \times bty\_avg\end{aligned}
\]

We can plot this line and the line corresponding to males with the following 
custom function.

```{r twoLines, eval = FALSE}
multiLines(m_bty_gen)
```

9.  What is the equation of the line corresponding to males? (*Hint:* For
    males, the parameter estimate is multiplied by 1.) For two professors
    who received the same beauty rating, which gender tends to have the
    higher course evaluation score?

The coefficient for gendermale is 0.17239 - since male is marked as 1, that means that on average males' beauty average was 0.17239 higher than females.

The decision to call the indicator variable `gendermale` instead of`genderfemale`
has no deeper meaning. R simply codes the category that comes first 
alphabetically as a $0$. (You can change the reference level of a categorical 
variable, which is the level that is coded as a 0, using the`relevel` function. 
Use `?relevel` to learn more.)

10. Create a new model called `m_bty_rank` with `gender` removed and `rank` 
    added in. How does R appear to handle categorical variables that have more 
    than two levels? Note that the rank variable has three levels: `teaching`, 
    `tenure track`, `tenured`.

```{r scatter-score-bty_avg_rank, eval = FALSE}
m_bty_rank <- lm(score ~ bty_avg + rank, data = evals)
summary(m_bty_rank)
```

R appears to handle having categorical variables with more than two levels by giving coefficients to all but one of the values. In this case ranktenure and ranktenured each get a coefficient, but not teaching.  Presumably for each of those variables, those are marked by 1 and teaching by 0.

The interpretation of the coefficients in multiple regression is slightly 
different from that of simple regression. The estimate for `bty_avg` reflects
how much higher a group of professors is expected to score if they have a beauty
rating that is one point higher *while holding all other variables constant*. In
this case, that translates into considering only professors of the same rank 
with `bty_avg` scores that are one point apart.

## The search for the best model

We will start with a full model that predicts professor score based on rank, 
ethnicity, gender, language of the university where they got their degree, age, 
proportion of students that filled out evaluations, class size, course level, 
number of professors, number of credits, average beauty rating, outfit, and 
picture color.

11. Which variable would you expect to have the highest p-value in this model? 
    Why? *Hint:* Think about which variable would you expect to not have any 
    association with the professor score.
    
My guess is that either outfit or picture color would have the highest p-value.

Let's run the model...

```{r m_full, eval = FALSE, tidy = FALSE}
m_full <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_profs + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full)
```

12. Check your suspicions from the previous exercise. Include the model output
    in your response.

Call:
lm(formula = score ~ rank + ethnicity + gender + language + age + 
    cls_perc_eval + cls_students + cls_level + cls_profs + cls_credits + 
    bty_avg + pic_outfit + pic_color, data = evals)

Residuals:
     Min       1Q   Median       3Q      Max 
-1.77397 -0.32432  0.09067  0.35183  0.95036 

Coefficients:
                        Estimate Std. Error t value Pr(>|t|)    
(Intercept)            4.0952141  0.2905277  14.096  < 2e-16 ***
ranktenure track      -0.1475932  0.0820671  -1.798  0.07278 .  
ranktenured           -0.0973378  0.0663296  -1.467  0.14295    
ethnicitynot minority  0.1234929  0.0786273   1.571  0.11698    
gendermale             0.2109481  0.0518230   4.071 5.54e-05 ***
languagenon-english   -0.2298112  0.1113754  -2.063  0.03965 *  
age                   -0.0090072  0.0031359  -2.872  0.00427 ** 
cls_perc_eval          0.0053272  0.0015393   3.461  0.00059 ***
cls_students           0.0004546  0.0003774   1.205  0.22896    
cls_levelupper         0.0605140  0.0575617   1.051  0.29369    
cls_profssingle       -0.0146619  0.0519885  -0.282  0.77806    
cls_creditsone credit  0.5020432  0.1159388   4.330 1.84e-05 ***
bty_avg                0.0400333  0.0175064   2.287  0.02267 *  
pic_outfitnot formal  -0.1126817  0.0738800  -1.525  0.12792    
pic_colorcolor        -0.2172630  0.0715021  -3.039  0.00252 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

It looks like I was wrong.  The highest p value appears to be for whether the class was taught by single or multiple professors, followed by the class level (lower or upper.)

13. Interpret the coefficient associated with the ethnicity variable.

There is no statistically significant association between the ethnicity of a professor and his/her score. For what it's worth, the coefficient is positive for non-minority professors, meaning overall non-minority professors tended to get higher scores, though again not in a statistically significant way.

14. Drop the variable with the highest p-value and re-fit the model. Did the
    coefficients and significance of the other explanatory variables change?
    (One of the things that makes multiple regression interesting is that
    coefficient estimates depend on the other variables that are included in
    the model.) If not, what does this say about whether or not the dropped
    variable was collinear with the other explanatory variables?

```{r m_full_rev, eval = FALSE, tidy = FALSE}
m_full_rev <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_students + cls_level + cls_credits + bty_avg 
             + pic_outfit + pic_color, data = evals)
summary(m_full_rev)
```

Residuals:
    Min      1Q  Median      3Q     Max 
-1.7836 -0.3257  0.0859  0.3513  0.9551 

Coefficients:
                        Estimate Std. Error t value Pr(>|t|)    
(Intercept)            4.0872523  0.2888562  14.150  < 2e-16 ***
ranktenure track      -0.1476746  0.0819824  -1.801 0.072327 .  
ranktenured           -0.0973829  0.0662614  -1.470 0.142349    
ethnicitynot minority  0.1274458  0.0772887   1.649 0.099856 .  
gendermale             0.2101231  0.0516873   4.065 5.66e-05 ***
languagenon-english   -0.2282894  0.1111305  -2.054 0.040530 *  
age                   -0.0089992  0.0031326  -2.873 0.004262 ** 
cls_perc_eval          0.0052888  0.0015317   3.453 0.000607 ***
cls_students           0.0004687  0.0003737   1.254 0.210384    
cls_levelupper         0.0606374  0.0575010   1.055 0.292200    
cls_creditsone credit  0.5061196  0.1149163   4.404 1.33e-05 ***
bty_avg                0.0398629  0.0174780   2.281 0.023032 *  
pic_outfitnot formal  -0.1083227  0.0721711  -1.501 0.134080    
pic_colorcolor        -0.2190527  0.0711469  -3.079 0.002205 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

Residual standard error: 0.4974 on 449 degrees of freedom
Multiple R-squared:  0.187,	Adjusted R-squared:  0.1634 
F-statistic: 7.943 on 13 and 449 DF,  p-value: 2.336e-14

No, the coefficients and p values did not change much. This suggests the dropped variable was collinear with the remaining variables and did not add much to the model on its own.

15. Using backward-selection and p-value as the selection criterion,
    determine the best model. You do not need to show all steps in your
    answer, just the output for the final model. Also, write out the linear
    model for predicting score based on the final model you settle on.

Below is the model which gave me the highest adjusted R-squared (0.1821)

```{r m_best, eval = FALSE, tidy = FALSE}
m_best <- lm(score ~ rank + ethnicity + gender + language + age + cls_perc_eval 
             + cls_credits + bty_avg + pic_outfit + pic_color+`bty_f1lower`+`bty_f1upper` +`bty_f2upper`,data = evals)
summary(m_best)
```

Expected score = 
4.087 +
(ranktenure track *  -0.184) +
(ranktenured * -0.0798) +
(ethnicitynot minority * 0.156) +
(gendermale * 0.255) +
(languagenon-english * -0.228) +  
(age * -0.009) +
(cls_perc_eval * 0.004) +  
(cls_creditsone credit * 0.503) +  
(bty_avg * -0.14) +
(pic_outfitnot formal * -0.1) +    
(pic_colorcolor * -0.205) +
(bty_f1lower * 0.039) +
(bty_f1upper * 0.08) +
(bty_f2upper * 0.058)

16. Verify that the conditions for this model are reasonable using diagnostic 
    plots.

```{r residuals-best, eval=FALSE}
plot(m_best$residuals)
hist(m_best$residuals)
qqnorm(m_best$residuals)
qqline(m_best$residuals)
```

The residuals appear to be normally distributed and to be fairly constant against their fitted values.

17. The original paper describes how these data were gathered by taking a
    sample of professors from the University of Texas at Austin and including 
    all courses that they have taught. Considering that each row represents a 
    course, could this new information have an impact on any of the conditions 
    of linear regression?

I'm not sure I understand the question. If we are talking abou increasing the number of observations, this would presumably help us improve the model and likely improve the normality of the observations and their residuals.

18. Based on your final model, describe the characteristics of a professor and 
    course at University of Texas at Austin that would be associated with a high
    evaluation score.
    
Characteristics of a professor and course associated with a high score: teaching professsor, non-minority ethnicity, male, english speaker, younger, high no and percent of students completing evaluation, lower average beauty score, pictured in black and white, and rated high on beauty by the upper class female students.

19. Would you be comfortable generalizing your conclusions to apply to professors

No, I would not feel comfortable generalizing my conclusions. While the sample was large, it was all from a single university in Texas. Further, it was not an experimental but instead an observational study.

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0). This lab was written by 
Mine &Ccedil;etinkaya-Rundel and Andrew Bray.
</div>