---
title: "Data Acquisition and Management (607) Assignment"
author: "Alex Low"
date: "September 11, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this assignment, we were tasked with creating a SQL database that captured the ratings of popular movies from friends or colleagues.

I started by creating some relational tables in MySQL - one for reviewers, one for movies, and one for movie reviews.  I then decided to try importing it into R using RODBC. Below is the code (user name and password blanked out for security reasons.)

```{r create-simpledfs-sql, eval=FALSE}
library(RODBC)
conn <- odbcConnect(dsn="alow-dsn", uid="xxxx", pwd="xxxx")
reviewerdf <- sqlQuery(conn, "SELECT * FROM Data607_Assignment2.Reviewers")
moviedf <- sqlQuery(conn, "SELECT * FROM Data607_Assignment2.Pop_Movies")
reviewdf <- sqlQuery(conn, "SELECT * FROM Data607_Assignment2.PopMovieReviews")
```

I also tried incorporating another SQL query which joined tables to show movies and reviewers in the same table.  Both seemed to work fine.

```{r create-detaileddfs-sql, eval=FALSE}
reviewdetaildf <- sqlQuery(conn, "SELECT Data607_Assignment2.Reviewers.Real_Name as Reviewer, Data607_Assignment2.Pop_Movies.Real_Name as Movie, PopMovieReviews.Rating as Rating
FROM Data607_Assignment2.Reviewers
LEFT JOIN Data607_Assignment2.PopMovieReviews
ON Reviewers.Person_ID = PopMovieReviews.Person_ID
LEFT JOIN Data607_Assignment2.Pop_Movies
ON Pop_Movies.Movie_ID = PopMovieReviews.Movie_ID
ORDER BY Data607_Assignment2.Pop_Movies.Real_Name, Data607_Assignment2.Reviewers.Real_Name")
View(reviewdetaildf)

```